---
title: "Statistical assignment 3"
author: "Simone Long_135288"
date: "15/02/2020"
output: github_document
---

```{r setup, include=FALSE}
# Please note these options.
# This tells R Markdown that we want to show code in the output document.
knitr::opts_chunk$set(echo = TRUE)
# Switching off messages in the output document. 
knitr::opts_chunk$set(message = FALSE)
# Switching on caching to make things faster (don't commit cache files on Github).
knitr::opts_chunk$set(cache = TRUE)
```

## Read data

```{r}
library(tidyverse)
library(data.table)

files <- dir(
             "C:/Users/simon/OneDrive/Documents/datan3_2019/data/UKDA-6614-tab/tab",
             pattern = "indresp",
             recursive = TRUE,
             full.names = TRUE)

files <- files[stringr::str_detect(files, "ukhls")]

vars <- c("memorig", "sex_dv", "age_dv", "vote6")

for (i in 1:7) {
        varsToSelect <- paste(letters[i], vars, sep = "_")
        varsToSelect <- c("pidp", varsToSelect)
        data <- fread(files[i], select = varsToSelect)
        if (i == 1) {
                all7 <- data  
        }
        else {
                all7 <- full_join(all7, data, by = "pidp")
        }
        rm(data)
} 

```

## Reshape data (20 points)

```{r}

Long <- all7 %>%
  gather(a_memorig:g_vote6, key = "variable", value = "value") %>%
  separate(variable, into = c("wave", "variable"), sep = "_", extra =  "merge") %>%
   spread(key = variable, value = value)

```

## Filter and recode (20 points)

```{r}
Long <- Long %>%
        filter(memorig == 1) %>%
        mutate(sex_dv = ifelse(sex_dv == 2, "female",
                           ifelse(sex_dv == 1, "male", NA))) %>%
        mutate(vote6 = recode(vote6,
                              `1` = 1,
                              `2` = 2,
                              `3` = 3,
                              `4` = 4,
                              .default = NA_real_))
        
  table(Long$sex_dv)
  table(Long$vote6)

```

## Calculate mean political interest by sex and wave (10 points)

```{r}
meanVote6 <- Long %>%
        group_by(wave, sex_dv) %>%
        filter(!is.na(sex_dv)) %>%
        summarise(voteavg = mean(vote6, na.rm = TRUE))
        
meanVote6
```

## Reshape the data frame with summary statistics (20 points)

```{r}
meanVote6 %>%
       spread(wave, voteavg)
```

Across all seven waves, women tend to be slightly less interested in politics than men on average. There are a variety of possible reasons for this. It could be that the sample of women for this dataset is uniquely disinterested in poitics. It could also be that the mean for interest in politics could be influenced in one direction based on age, rather than sex alone (i.e. maybe younger/older people tend to be more/less interested). But from a psychological perspective, it could be that women don't tend to engage as much politically because they don't see themselves represented in politics; there are far more male politicans than female, so it is conceivable that women don't think about politics as much because they don't feel they have a voice.

## Estimate stability of political interest (30 points)

```{r} 


Long <- Long %>%
  pivot_wider(id_cols = c(pidp, sex_dv), names_from = wave,
              values_from = c(age_dv, vote6))

Long <- Long %>%
  drop_na(vote6_a:vote6_g) %>%
  select(-c(age_dv_b:age_dv_g))

Long <- Long %>%
  mutate(delta = 
           abs(vote6_b - vote6_a) +
           abs(vote6_c - vote6_b) +
           abs(vote6_d - vote6_c) +
           abs(vote6_e - vote6_d) +
           abs(vote6_f - vote6_e) +
           abs(vote6_g - vote6_f))

sex.mean <- Long %>%
  group_by(sex_dv) %>%
  summarise(delta.mean = mean(delta))

knitr::kable(
  sex.mean,
 col.names = c("Sex", "Mean Delta")
)

age.mean <- Long %>%
  group_by(age_dv_a) %>%
  summarise(delta.mean = mean(delta))

knitr::kable(
  age.mean,
 col.names = c("Age", "Mean Delta")
)

library(ggplot2)
age.mean %>%
  ggplot(aes(x=age_dv_a, y=delta.mean, group=1)) +
  stat_smooth(method = 'loess', formula = y~x) +
  xlab("Age") +
  ylab("Mean Delta")
  
```

When looking at the mean change in political interest broken down by gender, it seems that men tend slightly more to vary in degree of interest than do women. This difference isn't incredibly significant, but could potentially be explained by such factors as female representation in politics influencing how much/little women participate and/or the increasing presence of news on social media causing people to engage more/less.

As far as age is concerned, there doesn't appear to be much variation in political interest. However, it does appear that those older than 85 do tend to alternate between greater and lesser engagement. This could quite simply be due to apathy brought on by old age, as the older one gets, the less they may feel they are affected by politics.